Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-18 appendix.tex
improved appendix b

Differentiating Eq. \ref{adia_iso1} properly gives
\begin{align}
  \left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
  \frac{d^2\delta v_z}{dz^2} + k_x^2c_s^2\left[\frac{\gamma}{\left(D + \gamma k_x^2 c_s^2\right)}\frac{dD}{dz} + \ii
    \frac{d\ln\rho}{dz}\left(\ii +
      \frac{q}{k_xr}\right)\right]\frac{d\delta v_z}{dz}\notag\\
  &+ \ii k_x^2 c_s^2 \left(\ii +
      \frac{q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} -
      \frac{1}{\left(D + \gamma
          k_x^2c_s^2\right)}\frac{dD}{dz}\frac{d\ln\rho}{dz}\right]\delta
    v_z. \label{adia_diso1}
\end{align}
The terms that were ignored in deriving Eq. \ref{adia_iso3} are those
proportional to $dD/dz$ in Eq. \ref{adia_diso1}. We estimate the
importance of these terms by noting that 
\begin{align}
  \frac{dD}{dz}\equiv \frac{d\kappa^2}{dz} \simeq \frac{d\Omega^2}{dz}
  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2},
\end{align}
for thin disks. Then Eq. \ref{adia_diso1} becomes
\begin{align}
\left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
  \frac{d^2\delta v_z}{dz^2} - k_x^2c_s^2\frac{d\ln\rho}{dz}\left[ 1 - 
      \frac{\ii q}{k_xr}  +  \frac{\gamma q c_s^2}{r^2\left(D +
        \gamma k_x^2 c_s^2\right)}\right]\frac{d\delta
    v_z}{dz}\notag\\ 
  &- k_x^2 c_s^2 \left(1 - 
      \frac{\ii q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} + 
      \frac{qc_s^2}{r^2\left(D + \gamma
          k_x^2c_s^2\right)}\left(\frac{d\ln\rho}{dz}\right)^2\right]\delta
    v_z. \label{adia_diso2}
\end{align}
Since $D\sim \Omega_k^2$ in the low-frequency limit, we 
see from Eq. \ref{adia_diso2} that the neglected terms are
$O(\epsilon^2)$. We can make this explicit by combining
Eq. \ref{adia_diso1}---\ref{adia_diso2} and Eq. \ref{adia_iso2}, then
set $D\to\Omega_k^2$. In terms of non-dimensional variables, the result
is  
\begin{align}
   &\delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon q\hat{k} -
    \frac{\gamma q \epsilon^2\hat{k}^2}{1+\gamma
      \hat{k}^2}\right)\ln\rho^{\prime}\delta v_z^\prime +
  \left[\left(\frac{1}{\gamma} + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \hat{k}^2\left(1 -
      \frac{\ii\epsilon
        q}{\hat{k}}\right)\left(\frac{\gamma-1}{\gamma} +
      \frac{q\epsilon^2}{1+\gamma\hat{k}^2}\right)\ln\rho^{\prime
      2}\right]\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{adia_iso3} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in
\S\ref{analytic_adia}.

>>>>

Differentiating Eq. \ref{adia_iso1} properly gives
\begin{align}
  \left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
  \frac{d^2\delta v_z}{dz^2} + k_x^2c_s^2\left[\frac{\gamma}{\left(D + \gamma k_x^2 c_s^2\right)}\frac{dD}{dz} + \ii
    \frac{d\ln\rho}{dz}\left(\ii +
      \frac{q}{k_xr}\right)\right]\frac{d\delta v_z}{dz}\notag\\
  &+ \ii k_x^2 c_s^2 \left(\ii +
      \frac{q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} -
      \frac{1}{\left(D + \gamma
          k_x^2c_s^2\right)}\frac{dD}{dz}\frac{d\ln\rho}{dz}\right]\delta
    v_z. \label{adia_diso1}
\end{align}
The terms that were ignored in deriving Eq. \ref{adia_iso3} are those
proportional to $dD/dz$ in Eq. \ref{adia_diso1}. Now, in the global
disk we have
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
  \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(2 + q +
    \frac{z^2/r^2-2}{z^2/r^2+1}\right). 
\end{align}
For the local problem we may then write
\begin{align}
  \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z;q),
\end{align}
where the function $F$ corresponds to the bracket in
Eq. \ref{dkappa2}. This function varies from $F=q$ at $z=0$ to $F\to
3+q$ as $|z|\to\infty$, i.e. its magnitude is of order unity. 
% We estimate the
% importance of these terms by noting that 
% \begin{align}
%   \frac{dD}{dz}\equiv \frac{d\kappa^2}{dz} \simeq \frac{d\Omega^2}{dz}
%   = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2},
% \end{align}
% for thin disks. 
Eq. \ref{adia_diso1} becomes
\begin{align}
\left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
  \frac{d^2\delta v_z}{dz^2} - k_x^2c_s^2\frac{d\ln\rho}{dz}\left[ 1 - 
      \frac{\ii q}{k_xr}  +  \frac{\gamma q c_s^2F}{r^2\left(D +
        \gamma k_x^2 c_s^2\right)}\right]\frac{d\delta
    v_z}{dz}\notag\\ 
  &- k_x^2 c_s^2 \left(1 - 
      \frac{\ii q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} + 
      \frac{qc_s^2F}{r^2\left(D + \gamma
          k_x^2c_s^2\right)}\left(\frac{d\ln\rho}{dz}\right)^2\right]\delta
    v_z. \label{adia_diso2}
\end{align}
Since $D\sim \Omega_k^2$ in the low-frequency limit, we 
see from Eq. \ref{adia_diso2} that the neglected terms are
$O(\epsilon^2)$. We can make this explicit by combining
Eq. \ref{adia_diso1}---\ref{adia_diso2} and Eq. \ref{adia_iso2}, then
set $D\to\Omega_k^2$. In terms of non-dimensional variables, the result
is  
\begin{align}
   &\delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon q\hat{k} -
    \frac{\gamma q \epsilon^2\hat{k}^2F}{1+\gamma
      \hat{k}^2}\right)\ln\rho^{\prime}\delta v_z^\prime +
  \left[\left(\frac{1}{\gamma} + \ii \epsilon q
      \hat{k}\right)\ln\rho^{\prime\prime} - \hat{k}^2\left(1 -
      \frac{\ii\epsilon
        q}{\hat{k}}\right)\left(\frac{\gamma-1}{\gamma} +
      \frac{q\epsilon^2F}{1+\gamma\hat{k}^2}\right)\ln\rho^{\prime
      2}\right]\delta v_z \notag\\&=
  -\hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right)\delta v_z\label{adia_diso3}.
\end{align}   
Eq. \ref{adia_diso3} differs from Eq. \ref{adia_iso3} by terms
proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
neglecting these terms has no qualitative effect on the discussion in
\S\ref{analytic_adia}.

### 2014-12-18 linear_iso.tex
added appendix b, edited first numerical stuff
improved appendix b

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. The problem simplifies because the eigenvalue now only
  appears in the last term in Eq. \ref{iso_ode}. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in Eq. \ref{iso_ode} through
  the factor $d\Omega^2/dz$. We therefore simplify further by ignoring
  the vertical dependence of $\kappa^2$ elsewhere. For clarity, we
  will replace $\kappa^2$ with $\Omega_k^2$.%, although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
These assumptions amount to making the replacement $D\to\Omega_k^2$.  
We will consider sound-speed profiles with $q<0$ in the global 
disk. Then $\Omega^2(z)$ decreases away from the midplane,
and may reach small values for sufficiently large $|z|$, which
invalidates the low-frequency approximation. In principle, then, the
above approximations implicitly limits us to consider thin disks ($\epsilon
\ll 1$) with vertical domain sizes not too large, so that the rotation
is nearly-Keplerian.

@@

\subsubsection{Polynomial solutions with vertical shear}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not neccessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\imag{\hat{\sigma}}>0$ is the growth rate and is
given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

>>>>

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. The problem simplifies because the eigenvalue now only
  appears in the last term in Eq. \ref{iso_ode}. We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in Eq. \ref{iso_ode} through
  the factor $d\Omega^2/dz$. We therefore simplify further by ignoring
  the vertical dependence of $\kappa^2$ elsewhere. For clarity, we
  will replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to the replacement
$D\to\Omega_k^2$.  We will consider sound-speed profiles with $q<0$ in the global 
disk. Then $\Omega^2(z)$ decreases away from the midplane,
and may reach small values for sufficiently large $|z|$, which
invalidates the low-frequency approximation. In principle, then, the
above approximations implicitly limits us to consider thin disks ($\epsilon
\ll 1$) with vertical domain sizes not too large, so that the rotation
is nearly-Keplerian.

@@

\subsubsection{Polynomial solutions with vertical shear}\label{iso_poly}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not neccessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\imag{\hat{\sigma}}>0$ is the growth rate and is
given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

### 2014-12-18 numerical.tex
improved appendix b

For this problem we solve Eq. \ref{iso_ode1}, using a pseudo-spectral
method by expanding $W$ in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equation on a grid with
$\zmax=10H_\mathrm{iso}$. This procedure converts the linear problem
to a standard matrix eigenvalue problem.

>>>>

For this problem we solve Eq. \ref{iso_ode}, with $D$ replaced by
$\kappa^2$, using a pseudo-spectral
method by expanding $W$ in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equation on a grid with
$\zmax=10H_\mathrm{iso}$. This procedure converts the linear problem
to a standard matrix eigenvalue problem.
